---
title: "ggvis cookbook"
output:
  html_document:
    toc: true
---

<!--
%\VignetteEngine{knitr}
%\VignetteIndexEntry{ggvis cookbook}
-->

```{r, echo = FALSE, message = FALSE}
library(knitr)
opts_chunk$set(comment = "#>", error = FALSE, tidy = FALSE)
opts_chunk$set(fig.width = 2, fig.height = 1.25, dpi = 100)
```

The following examples show you how to create a selection of common graphics with ggvis. First, load ggvis and dplyr:

```{r, message = FALSE}
library(ggvis)
library(dplyr)
```

## Scatterplots

We'll use the built-in `mtcars` data set, and look at two columns of interest, `mpg`, and `wt`:

```{r}
# The first few rows of mtcars
head(mtcars)
mtcars %>% ggvis(~wt, ~mpg) %>% layer_points()
```

Smaller points, a different shape, a different outline (stroke) color, and empty fill:

```{r}
mtcars %>% 
  ggvis(~wt, ~mpg) %>% 
  layer_points(size := 25, shape := "diamond", stroke := "red", fill := NA)
```


## Regression lines

Adding a smooth line

```{r, message = FALSE}
mtcars %>% 
  ggvis(~wt, ~mpg) %>%
  layer_points() %>%
  layer_smooths()
```

With a linear model, and 95% confidence interval for the model:

```{r, message = FALSE}
mtcars %>% 
  ggvis(~wt, ~mpg) %>%
  layer_points() %>%
  layer_model_predictions(model = "lm", se = TRUE)
```


## Scatter plots with grouping

Coloring points by a variable:

```{r, message = FALSE}
mtcars %>% 
  ggvis(~wt, ~mpg) %>% 
  layer_points(fill = ~factor(cyl))
```

Coloring points, and adding a smoother for each group. The grouping variable (which is applied before the `transform_smooth` is calculated) must be specified with `group_by()`:

```{r, message = FALSE}
mtcars %>% 
  ggvis(~wt, ~mpg, fill = ~factor(cyl)) %>% 
  layer_points() %>% 
  group_by(cyl) %>% 
  layer_model_predictions(model = "lm")
```

## Bar graphs

We'll use the built-in `pressure` data set for these examples:

```{r}
# The first few rows
head(pressure)
```

When the variable on the x axis is continuous (e.g., numeric or date-time):

```{r, message = FALSE, fig.width = 4}
pressure %>% 
  ggvis(~temperature, ~pressure) %>%
  layer_bars()
```

It's possible to specify the width of the bars:

```{r, message = FALSE, fig.width = 4}
pressure %>% 
  ggvis(~temperature, ~pressure) %>%
  layer_bars(width = 10)
```

When the variable on the x axis is categorical (e.g., factor or character):

```{r, message = FALSE, fig.width = 4}
# First, modify the pressure data set so that the x variable is a factor
pressure2 <- pressure %>% mutate(temperature = factor(temperature))

pressure2 %>% ggvis(~temperature, ~pressure) %>%
  layer_bars()
```

Notice that in this example, the x values aren't sorted quite right: they're sorted lexically, by the first, second, and then third digit, instead of by the numeric value. This is due to a limitation in ggvis.

It's possible to work around this by using a data set with a numerical x, and then tell it to display using a nominal scale. Notice in the example below, the x-axis labels are centered under each bar, each bar has a label, and they are sorted numerically.


```{r, message = FALSE, fig.width = 4}
# Use the original pressure data set with continuous x values
pressure %>% ggvis(~temperature, ~pressure) %>%
  layer_bars()
```


## Line graphs

```{r, message = FALSE}
pressure %>% ggvis(~temperature, ~pressure) %>% layer_lines()
```

Lines with points:

```{r, message = FALSE}
pressure %>% ggvis(~temperature, ~pressure) %>%
  layer_points() %>% 
  layer_lines()
```

## Histograms

We'll use the built-in `faithful` data set for these examples:

```{r}
# The first few rows
head(faithful)
```

Basic histogram:

```{r, message = FALSE}
faithful %>% ggvis(~eruptions) %>% layer_histograms()
```

Modify the fill color and binwidth, and add titles for the axes, since the automatic titles aren't very informative:

```{r, message = FALSE}
faithful %>% ggvis(~eruptions, fill := "#fff8dc") %>%
  layer_histograms(binwidth = 0.25) %>%
  add_axis("x", title = "eruptions") %>%
  add_axis("y", title = "count")
```
