---
title: "Data hierarchy"
---

<!--
%\VignetteEngine{knitr}
%\VignetteIndexEntry{Data hierarchy}
-->

```{r, echo = FALSE, message = FALSE}
library(knitr)
library(ggvis)
opts_chunk$set(comment = "#>", error = FALSE, tidy = FALSE)
```

Compared to ggplot2, ggvis has a much richer data hierarchy. In ggplot2, you
could define a data set and aesthetic mappings in the base plot, and override
them in each layer, but since layers could not contain other layers, there were
only ever two levels in the tree.  ggvis is more flexible because ggvis nodes
(the equivalent of ggplot2 layers) can contain child nodes. This makes it
possible to have whatever level of hierarchy that is best suited to your plot.

This is also related to a confusion in ggplot2 where `geom` often were actually
aliases for a geom + combination.  For example:

* `geom_histogram` = `geom_bar` + `stat_bin`
* `geom_smooth` = `geom_smooth` + `stat_smooth`

In ggvis, there is a clear demarkation between marks (pure drawing) and
layers (transformation + drawing). The low-level functions for drawing a mark start with `emit_`, and then the type of mark. For example:

* `layer_histograms` = `compute_bin` + `emit_rects`
* `layer_smooths` = `layer_model_predictions`  + (`emit_paths` + `emit_ribbons`)

ggplot2 needed special geoms like `geom_smooth` because the data hierarchy
was not deep enough, and there was no standard way for a stat to take the
output of another stat as it's input, and no way for a stat to feed data to
multiple layers without either creating a custom geom or duplicating the
stat multiple times.

## Data pipeline

A ggvis specification creates a data pipeline that flows from the starting
node to all of the leaves (marks). 

## Combining props

In ggplot2, layers had an `inherit.aes` property which control whether or
not a layer would inherit properties from the parent dataset - this is
particularly useful when writing functions that add annotation to arbitrary
plots - you don't want other properties that the user set interfering with your
layer.  In ggvis, that's now a property of `props()`: `props(inherit = FALSE)`.

To see how ggvis combines properties, you can use the `merge_props` function:

```{r}
merge_props <- ggvis:::merge_props
merge_props(props(x = ~x), props(y = ~y))
merge_props(props(x = ~a), props(x = ~b))
merge_props(props(x = ~a, y = ~a), props(x = ~b, inherit = FALSE))
```

There is currently no way to remove a prop inherited from the parent. See
https://github.com/rstudio/ggvis/issues/37 for progress.

## Case studies

### Minard's march

```{r, eval = FALSE}
ggplot(Minard.cities, aes(x = long, y = lat)) +
  geom_path(
    aes(size = survivors, colour = direction, group = group),
    data = Minard.troops
  ) +
  geom_point() +
  geom_text(aes(label = city), hjust=0, vjust=1, size=4)
```

In ggvis, we can make it a little more clear that we have one marked based on
survivors dataset and two marks based on the cities dataset.

```{r, eval = FALSE}
ggvis(data = NULL, x = ~long, y = ~lat) %>%
  layer_points(size = ~survivors, stroke = ~direction, data = Minard.troops) %>%
  layer_text(text := ~city, dx := 5, dy := -5, data = Minard.cities)
```
