<!--
%\VignetteEngine{knitr}
%\VignetteIndexEntry{Data hierarchy}
-->

```{r, echo = FALSE}
library(knitr)
opts_chunk$set(comment = "#", error = TRUE, tidy = FALSE)
```

# Data hierarchy
  
Compared to ggplot2, ggvis has a much richer data hierarchy. In ggplot2, you 
could define a data set and aesthetic mappings in the base plot, and override 
them in each layer, but since layers could not contain other layers, there were
only ever two levels in the tree.  ggvis is more flexible because ggvis nodes
(the equivalent of ggplot2 layers) can contain child nodes. This makes it 
possible to have whatever level of hierarchy that is best suited to your plot.

This is also related to a confusion in ggplot2 where `geom` often were actually
aliases for a geom + combination.  For example:

* `geom_histogram` = `geom_bar` + `stat_bin`
* `geom_smooth` = `geom_smooth` + `stat_smooth`

In ggvis, there is a clear demarkation between marks (pure drawing) and 
branches (drawing + transformation):

* `branch_histogram` = `mark_bar` + `transform_bin`
* `branch_smooth` = (`mark_line` + `mark_area`) + `transform_smooth`

ggplot2 needed special geoms like `geom_smooth` because the data hierarchy
was not deep enough, and there was no standard way for a stat to take the 
output of another stat as it's input, and no way for a stat to feed data to
multiple layers without either creating a custom geom or duplicating the 
stat multiple times.

## Data pipeline

A ggvis specification creates a data pipeline that flows from the starting
node to all of the leaves (marks). There are two ways to create a data 
pipeline:

* explicitly with pipeline: `pipeline(mtcars, transform_bin())`

* implicitly in ggvis/node: `ggvis(mtcars, transform_bin())`
  The data pipeline is stored in the data element of the gigvis_node object.

To see how a pipeline works, we'll create some explicitly and then flow 
`flow()` properties down them pipeline and see what happens.  
Since pipelines can also contain reactive data sources, this has to be done in
a reactive environment. To avoid this complication, we'll use `sluice()` which
sets up everything we need.

In this section we'll explore explicitly creating pipelines to help understand
what is going on behind the scenes. We'll start with a very simple pipeline that just contains a dataset. Sluicing it just returns the data as is. 

```{r}
df <- data.frame(x = 1:9, grp = rep(1:3, each = 3))

# A dataset has a name, and unique id (hash) in bracket
a <- pipeline(df)
a
sluice(a, props(x ~ x))
```

The next two pipelines add a data transformation. `b1` summarises the data 
by binning it. `b2` splits the data into pieces according to the value of the
`grp` variable. `b2` returns a special data structure called a `split_df`:
transforms need to be written specifically to understand how to operate with
this datastructure. Slucing the pipeline performs those transformations:

```{r}
b1 <- pipeline(df, transform_bin(binwidth = 3))
b1
b2 <- pipeline(df, by_group("grp"))
b2

sluice(b1, props(x ~ x))
sluice(b2, props(x ~ x))
```

Finally, `c` combines the operation of `b1` and `b2`, first splitting the data into pieces, then summarising each group with binned counts:

```{r}
c <- pipeline(df, by_group("grp"), transform_bin(binwidth = 3))
c
sluice(c, props(x ~ x))
```

Note that pipeline will automatically remove redundancies from a pipeline:

```{r}
pipeline(mtcars, sleep)
pipeline(mtcars, by_group("cyl"), sleep)
```

## Combining props

In ggplot2, layers had an `inherit.aes` property which control whether or
not a layer would inherit properties from the parent dataset - this is 
particularly useful when writing functions that add annotation to arbitrary
plots - you don't want other properties that the user set interfering with your
layer.  In ggvis, that's now a property of `props()`: `props(inherit = FALSE)`.

To see how ggvis combines properties, you can use the `merge_props` function:

```{r}
merge_props(props(x ~ x), props(y ~ y))
merge_props(props(x ~ a), props(x ~ b))
merge_props(props(x ~ a, y ~ a), props(x ~ b, inherit = FALSE))
```

There is currently no way to remove a prop inherited from the parent. See
https://github.com/rstudio/ggvis/issues/37 for progress.

## Case studies

### Minard's march

```R
ggplot(cities, aes(x = long, y = lat)) + 
  geom_path(
    aes(size = survivors, colour = direction, group = group), 
    data = troops
  ) + 
  geom_point() + 
  geom_text(aes(label = city), hjust=0, vjust=1, size=4)
```

In gigvis, we can make it a little more clear that we have one marked based on 
survivors dataset and two marks based on the cities dataset.

```R
ggvis(props(long, lat), 
  node(
    survivors, 
    props(size ~ survivors, colour ~ direction, group ~ group),
    mark_symbol()),
  node(
    cities,
    mark_symbol()
    mark_text(props(label ~ city))
  )
)
```
