---
title: ggvis basics
output:
  html_document:
    toc: true
runtime: shiny
---

<!--
%\VignetteEngine{knitr}
%\VignetteIndexEntry{ggvis basics}
-->

```{r, echo = FALSE, message = FALSE}
library(knitr)
library(ggvis)
opts_chunk$set(comment = "#>", error = FALSE, tidy = FALSE)
opts_chunk$set(fig.width = 3.5, fig.height = 2.5, dpi = 100)
```

## Introduction

The goal of ggvis is to make it easy to build interactive graphics for exploratory data analysis. ggvis has a similar underlying theory to [ggplot2](http://ggplot2.org) (the grammar of graphics), but it's expressed a little differently, and adds new features to make your plots interactive. ggvis also incorporates [shiny](http://www.rstudio.com/shiny/)'s reactive programming model and [dplyr](http://github.com/hadley/dplyr)'s grammar of data transformation.

The graphics produced by ggvis are fundamentally web graphics and work very differently from traditional R graphics. This allows us to implement exciting new features like interactivity, but it comes at a cost. For example, every interactive ggvis plot must be connected to a running R session (static plots do not need a running R session to be viewed). This is great for exploration, because you can do anything in your interactive plot you can do in R, but it's not so great for publication. We will overcome these issues in time, but for now be aware that we have many existing tools to reimplement before you can do everything with ggvis that you can do with base graphics.

This vignette is divided into four main sections:

* Dive into plotting with `ggvis()`.
* Add interactivity with the mouse and keyboard.
* Create more types of graphic by controlling the layer type.
* Build up rich graphics with multiple layers.

Each section will introduce to a major idea in ggvis, and point to more detailed explanation in other vignettes.

## `ggvis()`

Every ggvis graphic starts with a call to `ggvis()`. The first argument is the data set that you want to plot, and the other arguments describe how to map variables to visual properties. 

```{r}
p <- ggvis(mtcars, x = ~wt, y = ~mpg)
```

This doesn't actually plot anything because you haven't told ggvis how to display your data. You do that by layering visual elements, for example with `layer_points()`:

```{r}
layer_points(p)
```

(If you're not using RStudio, you'll notice that this plot opens in your web browser. That's because all ggvis graphics are web graphics, and need to be shown in the browser. RStudio includes a built-in browser so it can show you the plots directly.)

All ggvis functions take the visualisation as the first argument and return a modified visualisation. This seems a little bit awkward. Either you have to create temporary variables and modify them, or you have to use a lot of parentheses:

```{r}
layer_points(ggvis(mtcars, x = ~wt, y = ~mpg))
```

To make life easier ggvis uses the `%>%` (pronounced pipe) function from the magrittr package. That allows you to rewrite the previous function call as:

```{r}
mtcars %>%
  ggvis(x = ~wt, y = ~mpg) %>%
  layer_points()
```

Don't worry if this looks a little strange at first. You'll soon get used to it! This style of programming also allows gives you a lot of power when you start creating a lot of power, and allows you to seemlessly intermingle ggvis and dplyr code:

```{r, message = FALSE}
library(dplyr)
mtcars %>%
  ggvis(x = ~mpg, y = ~disp) %>%
  mutate(disp = disp / 61.0237) %>% # convert engine displacment to litres
  layer_points()
```    

The format of the visual properties needs a little explanation. We use `~` before the variable name to indicate that we don't want to literally use the value of the `mpg` variable (which doesn't exist), but instead we want we want to use the `mpg` variable inside in the dataset. This is a common pattern in ggvis: we'll always use formulas to refer to variables inside the dataset.

The first two arguments to `ggvis()` are usually the position, so by convention you can drop `x` and `y`:

```{r}
mtcars %>%
  ggvis(~mpg, ~disp) %>%
  layer_points()
```    

You can add more variables to the plot by mapping them to other visual properties like `fill`, `stroke`, `size` and `shape`.

```{r}
mtcars %>% ggvis(~mpg, ~disp, stroke = ~vs) %>% layer_points()
mtcars %>% ggvis(~mpg, ~disp, fill = ~vs) %>% layer_points()
mtcars %>% ggvis(~mpg, ~disp, size = ~vs) %>% layer_points()
mtcars %>% ggvis(~mpg, ~disp, shape = ~factor(cyl)) %>% layer_points()
```

If you want to make the points a fixed colour or size, you need to use `:=` instead of `=`. The `:=` operator means to use a raw, unscaled value. This seems like something that `ggvis()` should be able to figure out by itself, but making it explicit allows you to create some useful plots that you couldn't otherwise. See the [properties and scales](properties-scales.html) for more details.

```{r}
mtcars %>% ggvis(~wt, ~mpg, fill := "red", stroke := "black") %>% layer_points()
mtcars %>% ggvis(~wt, ~mpg, size := 300, opacity := 0.4) %>% layer_points()
mtcars %>% ggvis(~wt, ~mpg, shape := "cross") %>% layer_points()
```

## Interaction

As well as mapping visual properties to variables or setting them to specific values, you can also connect them to interactive controls.

The following example allows you to control the size and opacity of points with two sliders:

```{r}
mtcars %>% 
  ggvis(~wt, ~mpg, 
    size := input_slider(10, 100),
    opacity := input_slider(0, 1)
  ) %>% 
  layer_points()
```

You can also connect interactive components to other plot parameters like the width 
and centers of histogram bins:

```{r}
mtcars %>% 
  ggvis(~wt) %>% 
  layer_histograms(width =  input_slider(0, 2, step = 0.10, label = "width"),
                   center = input_slider(0, 2, step = 0.05, label = "center"))
```

Behind the scenes, interactive plots are built with [shiny](http://www.rstudio.com/shiny/), and you can currently only have one running at a time in a given R session. To finish with a plot, press the stop button in Rstudio, or close the browser window and then press Escape or Ctrl + C in R.

As well as `input_slider()`, ggvis provides `input_checkbox()`, `input_checkboxgroup()`, `input_numeric()`, `input_radiobuttons()`, `input_select()` and  `input_text()`. See the examples in the documentation for how you might use each one.

You can also use keyboard controls with `left_right()` and `up_down()`. Press the left and right arrows to control the size of the points in the next example.

```{r}
keys_s <- left_right(10, 1000, step = 50)
mtcars %>% ggvis(~wt, ~mpg, size := keys_s, opacity := 0.5) %>% layer_points()
```

You can also add on more complex types of interaction like tooltips:

```{r}
mtcars %>% ggvis(~wt, ~mpg) %>% 
  layer_points() %>% 
  add_tooltip(function(df) df$wt)
```

You'll learn more about complex interaction in the [interactivity](interactivity.html) vignette.

## Layers

So far, you seen two layer functions: `layer_points()` and `layer_histograms()`. There are many other layers, and they can be roughly categorised into two types:

* Simple, which include primitives like points, lines and rectangles.

* Compound, which combine data transformations with one or more simple layers.

All layer functions use the plural, not the singular. Think the verb, not the noun: I'm going to layer some points onto my plot.

### Simple layers

There are five simple layers:

* Points, `layer_points()`, with properties `x`, `y`, `shape`, `stroke`, 
  `fill`, `strokeOpacity`, `fillOpacity`, and `opacity`.

    ```{r}
    mtcars %>% ggvis(~wt, ~mpg) %>% layer_points()
    ```

* Paths and polygons, `layer_paths()`.

    ```{r}
    df <- data.frame(x = 1:10, y = runif(10))
    df %>% ggvis(~x, ~y) %>% layer_paths()
    ```

    If you supply a `fill`, you'll get a polygon

    ```{r}
    t <- seq(0, 2 * pi, length = 100)
    df <- data.frame(x = sin(t), y = cos(t))
    df %>% ggvis(~x, ~y) %>% layer_paths(fill := "red")
    ```

* Filled areas, `layer_ribbons()`. Use properties `y` and `y2` to control the
  extent of the area.

    ```{r}
    df <- data.frame(x = 1:10, y = runif(10))
    df %>% ggvis(~x, ~y) %>% layer_ribbons()
    df %>% ggvis(~x, ~y + 0.1, y2 = ~y - 0.1) %>% layer_ribbons()
    ```

* Rectangles, `layer_rects()`. The location and size of the rectangle is
  controlled by the `x`, `x2`, `y` and `y2` properties.

    ```{r}
    set.seed(1014)
    df <- data.frame(x1 = runif(5), x2 = runif(5), y1 = runif(5), y2 = runif(5))
    df %>% ggvis(~x1, ~y1, x2 = ~x2, y2 = ~y2, fillOpacity := 0.1) %>% layer_rects()
    ```

* Text, `layer_text()`. The text layer has many new options to control the
  apperance of the text: `text` (the label), `dx` and `dy` (margin in pixels
  between text and anchor point), `angle` (rotate the text), `font` (font
  name), `fontSize` (size in pixels), `fontWeight` (e.g. bold or normal),
  `fontStyle` (e.g. italic or normal.)

    ```{r}
    df <- data.frame(x = 3:1, y = c(1, 3, 2), label = c("a", "b", "c"))
    df %>% ggvis(~x, ~y, text := ~label) %>% layer_text()
    df %>% ggvis(~x, ~y, text := ~label) %>% layer_text(fontSize := 50)
    df %>% ggvis(~x, ~y, text := ~label) %>% layer_text(angle := 45)
    ```

### Compound layers

The four most common compound layers are:

* `layer_lines()` which automatically orders by the x variable:

    ```{r}
    t <- seq(0, 2 * pi, length = 20)
    df <- data.frame(x = sin(t), y = cos(t))
    df %>% ggvis(~x, ~y) %>% layer_paths()
    df %>% ggvis(~x, ~y) %>% layer_lines()
    ```
    
    `layer_lines()` is equivalent to `arrange()` + `layer_paths()`:
    
    ```{r}
    df %>% ggvis(~x, ~y) %>% arrange(x) %>% layer_paths()
    ```

* `layer_histograms()` and `layer_freqpolys()` which allows you to explore the 
  distribution of continuous. Both layers first bin the data with 
  `compute_bin()` then display the results with either rects or lines.
    
    ```{r}
    mtcars %>% ggvis(~mpg) %>% layer_histograms()
    
    # Or equivalently
    binned <- mtcars %>% compute_bin(~mpg) 
    binned %>% 
      ggvis(x = ~xmin_, x2 = ~xmax_, y2 = 0, y = ~count_, fill := "black") %>%
      layer_rects()
    ```

* `layer_smooths()` fits a smooth model to the data, and displays predictions
  with a line. It's used to highlight the trend in noisy data:

    ```{r}
    mtcars %>% ggvis(~wt, ~mpg) %>% layer_smooths()

    # Or equivalently
    smoothed <- mtcars %>% compute_smooth(mpg ~ wt)
    smoothed %>% ggvis(~pred_, ~resp_) %>% layer_paths()
    ```

    You can control the degree of wiggliness with the span parameter:

    ```{r}
    span <- input_slider(0.2, 1, value = 0.75)
    mtcars %>% ggvis(~wt, ~mpg) %>% layer_smooths(span = span)
    ```

You can learn more about layers in the [layers](layers.html) vignette.

## Multiple layers

Rich graphics can be created by combining multiple layers on the same plot. This easier to do: just layer on multiple elements:

```{r}
mtcars %>% 
  ggvis(~wt, ~mpg) %>% 
  layer_smooths() %>% 
  layer_points()
```

You could use this approach to add two smoothers with varying degrees of wiggliness:

```{r}
mtcars %>% ggvis(~wt, ~mpg) %>%
  layer_smooths(span = 1) %>%
  layer_smooths(span = 0.3, stroke := "red")
```

You'll learn more about building up rich hierarchical graphics in [data hierarchy](data-hierarchy.html).

## More details

There are other optional components that you can include:

* `scales`, to control the mapping between data and visual properties.
  These are described in the [properties and scales](properties-scales.html) vignette.

* `legends` and `axes` to control the appearance of the guides produced by
  the scales.  See the [axes and legends](axes-legends.html) vignette for more details.
