<!--
%\VignetteEngine{knitr}
%\VignetteIndexEntry{ggvis basics}
-->

```{r, echo = FALSE}
library(knitr)
library(ggvis)
opts_chunk$set(comment = "#", error = TRUE, tidy = FALSE, eval = FALSE)
```

# ggvis basics

Every ggvis graphic starts with a call to `ggvis()`. Every ggvis plot must 
contain one or more:

* data sets
* property mapping
* mark

A basic plot looks like this:

```{r}
ggvis(mtcars, props(x ~ wt, y ~ mpg), mark_symbol())
```

The order of the arguments doesn't matter, so the following specifications 
will create the same plot. However, it's customary to arrange arguments in the 
order: data, prop, mark.

```{r}
ggvis(mark_symbol(), mtcars, props(x ~ wt, y ~ mpg))
ggvis(props(x ~ wt, y ~ mpg), mtcars, mark_symbol())
```

There are also other optional components that you can include:

* `scales`, to control the mapping between data and visual properties. 
  These are described in the "properties and scales" vignette.

* `legends` and `axes` to control the appearance of the guides produced by
  the scales.  See the "axes and legends" vignette for more details.
  
* `nodes` to create a richer hierarchy of data and marks. See the "data 
  hierarchy" vignette to learn how this works.

Other vignettes describe:

* how to add __interactivity__, using basic interactive inputs or by embedding
  in a shiny app
  
* the full details of the __marks__

* how ggvis compares to __ggplot2__

* __vega__, the rendering engine used by ggvis
