<!--
%\VignetteEngine{knitr}
%\VignetteIndexEntry{ggvis basics}
-->

```{r, echo = FALSE}
library(knitr)
suppressPackageStartupMessages(library(ggvis))
opts_chunk$set(comment = "#", error = TRUE, tidy = FALSE, eval = FALSE)
```

# ggvis basics

Every ggvis graphic starts with a call to `ggvis()`. Every ggvis plot must 
contain one or more:

* data sets
* property mapping
* mark

A basic plot looks like this:

```{r}
ggvis(mtcars, props(x = ~ wt, y = ~ mpg), mark_symbol())
```

The order of the arguments doesn't matter, so the following specifications 
will create the same plot. However, it's customary to arrange arguments in the 
order: data, prop, mark.

```{r}
ggvis(mark_symbol(), mtcars, props(x = ~ wt, y = ~ mpg))
ggvis(props(x = ~ wt, y = ~ mpg), mtcars, mark_symbol())
```

## Data sets

You bring the data and ggvis brings the plots. Unlike ggvis, ggplot2 will eventually support multiple types of data, not just data frames.

You'll can also use transformation functions which take a dataset and modify it in some way.  See `transform_bin()` and `transform_smooth()` for examples.

## Property mappings

The first thing you really need to understand is how to map variables in your data to visual properties. In ggvis, you do that with `props()`, analogous to `aes()` in ggplot2. `props()` takes a sequence of name-value pairs:

* `props(x = ~wt, y = ~mpg)`: map x position to the `wt` variable, y position to `mpg`

* `props(x = ~factor(cyl), y = mpg / wt)`: map x position to `cyl` transformed to be a factor, and y to `mpg` standardised by `wt`

Note the use of `~`: we're using to refer to variables in the dataset, not variables in the local environment.

Unlike ggplot2, you also use `props()` to set constant values:

* `props(x = ~wt, y = ~mpg, stroke := "red", fill := "blue")`: as well as mapping `wt` to x position and `mpg` to y position, also _set_ the stroke to red and the fill to blue.

Note that to set values we use `:=` and not `=`.  This is because these are _raw_ values, values that are meaningful to the display system, not _data_ values, values that make sense to you. 

This gives us four combinations:

* `x = ~mpg`: map a variable
* `x = 1`: map a constant value 
* `x := ~mpg`: set a variable
* `x := 1`: set a constant

Most of the time you'll use either `x = ~mpg` or `fill := "red"`. But there are cases where the other options are useful, as you'll see in other vignettes.

## Marks

Marks do the actual drawing. 

* `mark_symbol`, for drawing points (like `geom_point`)

* `mark_line`, for drawing general paths and polygons (like `geom_path`, or if ordered, `geom_line`)

* `mark_rect`, for drawing rectangles (like `geom_rect`, `geom_bar`, `geom_tile`)

* `mark_area`, for filled areas (like `geom_area` and `geom_ribbon`)

* `mark_text`, for text (like `geom_text`)

* `mark_arc`, for drawing circular arcs

The basic properties are:

 * `x`, the x-coordinate, `x2` and `width` for marks with horizontal dimension
 * `y`, the y-coordinate, `y2` and `height` for marks with vertical dimension
 * `stroke` and `fill` colours
 * `opacity` of both stroke and fill
 
`mark_symbol` also understands:
 
 * `size`: the area of the symbol (in pixels)
 * `shape`: the shape of the symbol

Plus a quite few more that only apply to `mark_text` or `mark_arc`. See `?props` for a full list. 

## More details

There are also other optional components that you can include:

* `scales`, to control the mapping between data and visual properties. 
  These are described in the "properties and scales" vignette.

* `legends` and `axes` to control the appearance of the guides produced by
  the scales.  See the "axes and legends" vignette for more details.
  
* `nodes` to create a richer hierarchy of data and marks. See the "data 
  hierarchy" vignette to learn how this works.

Other vignettes describe:

* how to add __interactivity__, using basic interactive inputs or by embedding
  in a shiny app
  
* the full details of the __marks__

* how ggvis compares to __ggplot2__

* __vega__, the rendering engine used by ggvis
