---
title: Interactivity
output:
  html_document:
    toc: true
runtime: shiny
---

<!--
%\VignetteEngine{knitr}
%\VignetteIndexEntry{Interactivity}
-->

```{r, echo = FALSE, message = FALSE}
library(knitr)
library(ggvis)
opts_chunk$set(comment = "#>", error = FALSE, tidy = FALSE)
opts_chunk$set(fig.width = 3.5, fig.height = 2.5, dpi = 100)
```

**Note: If you're viewing the HTML version of this document generated with knitr, the examples will have their interactive features disabled. You'll need to run the code in R to see and use the interactive controls.**

Ggvis interactivity is built on top of Shiny's reactive programming
model. It's possible to use ggvis without understanding how Shiny works,
but you'll be limited to relatively simple interactions. The first part of
this vignette describes __basic interactive controls__, which provide a quick
and easy way to add basic interactivity to a plot. They are not very flexible,
but they cover the most common interactive needs and you don't need to know
anything about Shiny to use them.

The second part of this vignette shows you how to __embedding a ggvis in a Shiny
app__.  This gives you complete freedom to make any component of the plot
interactive, to display multiple plots on one page, and to freely arrange
controls. The downside is that you'll need more code, and you'll need at
least a basic understanding of Shiny. If you're not already familiar with
how Shiny works, I recommend that you first read through the
[Shiny tutorial](http://rstudio.github.io/shiny/tutorial/)

## Basic interactive controls

A plot with basic interactive controls looks very similar to a static plot.
You just replace constant values with functions that produce interactive
controls like `input_slider()`:

```{r}
mtcars %>%
  ggvis(~wt, ~mpg) %>%
  layer_smooths(span = input_slider(0.5, 1, value = 1)) %>%
  layer_points(size := input_slider(100, 1000, value = 100))
```

As well as `input_slider()`, which produces a slider (or a double-ended range
slider), there are a number of other interactive controls:

* `input_checkbox()`: a check-box
* `input_checkboxgroup()`: a group of check boxes
* `input_numeric()`: a spin box
* `input_radiobuttons()`: pick one from a set options
* `input_select()`: create a drop-down text box
* `input_text()`: arbitrary text input

Note that all interactive inputs start with `input_` so that you can always use
tab completion to remind you of the options.

Here is an example of a plot that uses both a slider and a select box:

```{r, eval=FALSE}
mtcars %>% ggvis(x = ~wt) %>%
    layer_densities(
      adjust = input_slider(.1, 2, value = 1, step = .1, label = "Bandwidth adjustment"),
      kernel = input_select(
        c("Gaussian" = "gaussian",
          "Epanechnikov" = "epanechnikov",
          "Rectangular" = "rectangular",
          "Triangular" = "triangular",
          "Biweight" = "biweight",
          "Cosine" = "cosine",
          "Optcosine" = "optcosine"),
        label = "Kernel")
    )
```

### Arguments

The arguments differ slightly depending on the type of input, and are documented
in the individual function documentation. All input sliders share three common
arguments:

* `label`: an informative text label displayed next to the control

* `id`: a unique identifier for the control - this is randomly assigned and
   you should only need to override in exceptional circumstances.

* `map`: a function that enables you to remap the output of the input to
  something appropriate for your graphic.

`map` is the most commonly argument used as it allows you to translate the value
recieved from the input to something that's meaningful for your plot. For
example, many parameters of statistical models are best expressed on a log
scale. You can use `map` in conjunction with an slider to make multiple orders
of magnitude easily accessible from a single slider.

```{r, eval = FALSE}
input_slider(-5, 5, label = "lambda", map = function(x) 10 ^ x)
```

### Multiple outputs

There are two ways to have an interactive input control multiple settings on
the plot.

1.  Use the regular property inheritance structure:

    ```{r}
    mtcars %>%
      ggvis(~wt, ~mpg, size := input_slider(10, 1000)) %>%
      layer_points(fill := "red") %>%
      layer_points(stroke := "black", fill := NA)
    ```

2. Create the input and assign it to a variable:

    ```{r}
    slider <- input_slider(10, 1000)
    mtcars %>% ggvis(~wt, ~mpg) %>%
      layer_points(fill := "red", size := slider) %>%
      layer_points(stroke := "black", fill := NA, size := slider)
    ```

    This is a little more typing, but is more flexible.

Compare the following two plots. In the first plot, one slider is created and
used for both layers - it controls the size of both the red and the black
points. In the second plot, you get two independent sliders which allow you to
control the size of the red and black points independently.

```{r}
slider <- input_slider(100, 1000)
mtcars %>% ggvis(~wt, ~mpg) %>%
  layer_points(size := slider) %>% 
  layer_points(fill := "red", opacity := 0.5, size := slider)

mtcars %>% ggvis(~wt, ~mpg) %>%
  layer_points(size := input_slider(100, 1000, label = "black")) %>%
  layer_points(fill := "red", size := input_slider(100, 1000, label = "red"))
```

### Property mappings

When you map a slider to a property, you have four basic options:

* constant, unscaled
* constant, scaled
* variable, unscaled
* variable, scaled

If you just assign the input in `props()`, like `props(size = input_slider())`,
you'll get a constant, unscaled property. To gain more control you can use
`prop()`: `props(prop("size", input_slider()))`.

You'll need to do this if you want to:

* select a variable to plot based on a dropdown:
  `prop(x = input_select(c("disp", "wt")), constant = FALSE)`

* annotate the plot in the data space:
  `prop(x = input_slider(min(mpg$wt), max(mpg$wt)), scale = TRUE)`

You can read more about these options in the __properties and scales__ vignette.

### Limitations

Currently, interactive inputs can only be used in two places:

* as arguments to transforms: `layer_smooths(span = input_slider(0, 1))`
* as properties: `props(size = input_slider(10, 1000))`

This means that interactive inputs can only modify the data, not the underlying
plot specification. In other words, with only basic interactivity there's no
way to add or remove layers, or switch between different datasets. This is a
reasonable limitation because if you're doing exploration you can always create
a new ggvis with R code, or if you're polishing a plot for presentation, you
can embed it in a Shiny app and gain full control over the plot.

### Compared to Shiny

If you're familiar with Shiny, you'll notice that these functions have very
similar equivalents: `sliderInput()`, `selectInput()` and so on.  There are two
main differences:

* the argument order has been tweaked so that you can create a basic control
  with minimal arguments. The `label` is optional because if you're creating a
  plot for yourself, you usually remember what the controls do; and each control
  is assigned a random `id` so you don't need to think one up.

* interactive input are not necessarily created in a reactive context, so they
  can not return reactives. Instead, they return delayed reactives, which are
  activated and connected together when the plot is displayed.

## Shiny apps

If you know how to create a Shiny app already, adding a ggvis plot is easy. In
`server.r`, create a ggvis object, and call `bind_shiny()`. When called in a pipe, the arguments are the names of the plot and control placeholders.

```{r, eval = FALSE}
# A reactive expression wrapper for input$size
input_size <- reactive(input$size)

mtcars %>% 
  ggvis(~disp, ~mpg, size := input_size) %>%
  layer_points() %>%
  bind_shiny("ggvis", "ggvis_ui")
```

Create `ui.r` as usual, using `ggvisOutput()` to place the ggvis plot on the
page.

```{r, eval = FALSE}
shinyUI(pageWithSidebar(
  sidebarPanel(
    sliderInput("size", "Area", 10, 1000),
  ),
  mainPanel(
    uiOutput("ggvis_ui"),
    ggvisOutput("ggvis")
  )
))
```

There are a couple of examples in the `demo/apps` directory.
