---
output:
  ggvis:::html_document
---

<!--
%\VignetteEngine{knitr}
%\VignetteIndexEntry{Marks}
-->

```{r, echo = FALSE, message = FALSE}
library(knitr)
library(ggvis)
opts_chunk$set(comment = "#>", error = FALSE, tidy = FALSE)
opts_chunk$set(fig.width = 3.5, fig.height = 2.5, dpi = 100)
```

# Layers

In ggvis, there are two types of layers. Simple layers correspond directly to vega marks and represent geometric primitives like points, lines and rectangles. Compound layers combine data transformations with one or more simple layers.

## Marks

Marks are the fundamental building block. Each type of mark represents a different type of geometric object, and so in some sense they are the equivalent of geoms in ggplot2. However, there are a far fewer marks than geoms because marks tend to be more capable, and because they never perform any statistical transformation.

Most of the time you don't use marks directly - you instead use their layer wrappers. That's because most of the time you don't want to think about the difference between simple and compound layers - you just want to make a plot. But if you want to understand more deeply how things work, or create your own layer functions, it's important to understand the distinction.

`ggvis` exposes
[five vega marks](https://github.com/trifacta/vega/wiki/Marks#color-references)
that have very similar ggplot2 equivalents:

* `mark_area()` = `geom_ribbon()`, `geom_area()`
* `mark_path()` = `geom_path()`, `geom_line()`
* `mark_rect()` = `geom_rect()`, `geom_bar()`, `geom_tile()`
* `mark_point()` = `geom_point()`
* `mark_text()` = `geom_text()`

There are two marks that have no equivalent in ggplot2:

* `mark_arc()` allows you to draw arcs. You could accomplish something
  similar in ggplot2 with `geom_rect()` with `coord_polar()` but it was
  relatively inflexible and required contorting your brain a bit.

* `mark_image()` allows you to place arbitrary images on a plot. There is
  currently little additional support for this in ggvis, so you'll need to
  make sure the images are publicly available somewhere. If you've found
  a compelling use-case for `layer_image()`, let us know so we can consider
  how to better support it.

And two vega marks that are not currently implemented in ggvis:

* Support for the group mark will be added in a future major release that
  will add support for facetting/subplots.

* I can't currently see any use for the path mark, which takes svg path
  specifications as input.

All marks take two optional arguments which are only needed if you want to
override the values inherited from parent:

* `props`, a `props()` object used to add any custom override for this layer
* `data`, an optional dataset to override the data inherited from the parent

## Adding a layer to a plot

For example, the following two plots are equivalent. In the first one, the
data and props are defined at the top level and inherited by the mark, and in
the second, they are defined in the mark itself. Note that the order of the
arguments to `ggvis()` and the mark function are different: this is because
you'll usually set the data in top-level plot object, and usually only override
props in the mark.

```{r, results = 'asis'}
ggvis(mtcars, props(x = ~wt, y = ~mpg, stroke := "red")) +
  layer_point()

ggvis() +
  layer_point(data = mtcars, props(x = ~wt, y = ~mpg, stroke := "red"))
```

You can add layers on to plots created by `qvis()`, but we don't recommend it if you're setting custom props. Using two different ways of describes props makes the code harder to read.

```{r, results = 'asis'}
qvis(mtcars, ~wt, ~mpg) +
  layer_smooth(props(stroke := "red"))

# This is only a little more typing but more symmetric
ggvis(mtcars, props(~wt, ~mpg)) +
  layer_point() +
  layer_smooth(props(stroke := "red"))
```

## Mark details

The following sections describe particular details related to individual marks.

### `layer_rect`

Must set two of `x`, `x2`, and `width`, and two of `y`, `y2` and `height`.

If you have an ordinal scale, set width and/or height to `prop_band()` to
occupy the complete band corresponding to that categorical value.

```{r, results = 'asis', eval = FALSE}
hec <- as.data.frame(xtabs(Freq ~ Hair + Eye, HairEyeColor))

ggvis(hec, props(x = ~Hair, y = ~Eye, fill = ~Freq),
  layer_rect(props(width := band(), y := band())))
```

### `layer_path`

`layer_path` produces lines if fill is empty, and polygons if it is set to a
value.

```{r, results = 'asis'}
df <- data.frame(x = c(1, 1, 2, 2), y = c(2, 1, 1, 2))
ggvis(df, props(x = ~x, y = ~y, stroke := "red"), layer_path())
# Add a fill colour to make it a polygon
ggvis(df, props(x = ~x, y = ~y, fill := "red"), layer_path())
```

You can not currently set the component of lines to different colours:
track progress at https://github.com/trifacta/vega/issues/122.

To create a line that spans the complete range of the plot, use `prop_group()`:

```{r, results = 'asis', eval = FALSE}
ggvis() +
  layer_point(props(x = ~disp, y = ~wt), mtcars) +
  layer_path(props(y = ~mean(mtcars$wt), x := 0, x2 := prop_group())) +
  layer_path(props(x = ~mean(mtcars$disp), y := 0, y2 := prop_group()))
```

Note that this won't work because `prop_group()` hasn't been written yet: track
progress at https://github.com/rstudio/ggvis/issues/38.

### `layer_area`

You need to set two of `y`, `y2` and `height`:

```{r, results = 'asis'}
df <- data.frame(x = 1:10, y = (1:10) ^ 2)
ggvis(df, props(x = ~x, y = ~y, y2 := 0), layer_area())

# Set height in pixels
ggvis(df, props(x = ~x, y = ~y, height := 20), layer_area())
```

Height can only be mapped to a constant, because it does not have an obvious
scale associated with it. You could force height to use the `y` scale, but that
doesn't work - the area hangs below the y line, and increasing the value of
height makes the area narrower! What's going on is that the underlying graphics
device has (0, 0) in the top-left corner, and so the y-scale is upside down. As
you increase height, it's mapped like a y variable so bigger values are further
away.

```{r, results = 'asis'}
ggvis(df, props(x = ~x, y = ~y, height := prop(80, scale = "y"))) +
  layer_area()

df <- data.frame(x = 1:10, y = (1:10) ^ 2)
ggvis(df, props(x = ~x, y = ~y),
  layer_area(props(height = prop(input_slider(0, 100), scale = "y"))),
  layer_path(props(stroke := "red", strokeWidth := 10))
)
```

If you want to set the height in data units, you need to
compute the offsets yourself:

```{r, results = 'asis'}
ggvis(df, props(x = ~x, y = ~y - 2, y2 = ~y + 2), layer_area())
```

The distinction between dimension and location is important in other graphics system, and a grid vignette `vignette("locndimn", package = "grid")` has a little more information about how things work in grid.

## Grouping

In ggplot2, grouping was a special aesthetic; in ggvis, grouping is a data
transformation: use `by_group()` to split your data up into pieces given a
specified variable, or `auto_split()` to split up by any categorical variable included in the plot:

```{r, results = 'asis'}
ggvis(mtcars, props(~wt, ~mpg)) +
  layer_point() +
  layer_path(data = by_group(cyl))

ggvis(mtcars, props(~wt, ~mpg, stroke = ~factor(cyl))) +
  layer_point() +
  layer_path(data = auto_split())
```

Some layers, like `layer_line()`, include `auto_split()` so will split automatically:

```{r, results = 'asis'}
ggvis(mtcars, props(~wt, ~mpg, stroke = ~factor(cyl))) +
  layer_line()
```

## Compound layers

Simple layers map directly to vega's primitive drawing functions (marks). Compound layers combine data transformation and simple layers (potentially multiple of each) to more sophisticated graphics. For example:

* `layer_histogram()` = `transform_bin()` + `mark_rect()`
* `layer_freqpoly()` = `transform_bin()` + `mark_path()`
* `layer_smooth()` = `transform_smooth()` + (`mark_path()` + `mark_area()`)

All layers start with `layer_`. The source is usually relatively simple - the complexity generally comes when figuring out how to map properties supplied to the layer to individual marks:

```{r}
layer_smooth
layer_histogram
```

If you find yourself using a set of marks commonly, it might be worth writing
your own layer function. Typically, the first argument should be `props` and
`...` should be passed on to the tranform function.

## Conversion from ggplot2

The following list provides a conversion from ggplot2 geoms to vega marks.
However, because ggvis currently provides few transformations, many translations
don't currently exists - but they are on the roadmap and will be added over
time.

* `geom_abline`
* `geom_area` = `layer_area` with `props(y ~ 0)`
* `geom_bar` = `layer_rect` with `props(y ~ 0, width = band())`
* `geom_bin2d` = `layer_rect` + `transform_bin2d`
* `geom_blank`
* `geom_boxplot`
* `geom_contour`
* `geom_crossbar`
* `geom_density` = `layer_path` + `transform_density`
* `geom_density2d`
* `geom_dotplot`
* `geom_errorbar`
* `geom_errorbarh`
* `geom_freqpoly`
* `geom_hex`
* `geom_histogram` = `layer_histogram` = `layer_bar` + `transform_bin`
* `geom_hline` =  = `layer_path` + `props(x2 ~ proo_group("width"))`.
* `geom_jitter`
* `geom_line` = `layer_path` + `transform_sort`
* `geom_linerange`
* `geom_map`
* `geom_path` = `layer_path`
* `geom_point` = `layer_point`
* `geom_pointrange`
* `geom_polygon` = `layer_path`
* `geom_quantile` = `layer_path` + `transform_quantreg`
* `geom_raster`
* `geom_rect` = `layer_rect`
* `geom_ribbon` = `layer_area`
* `geom_rug`
* `geom_segment`
* `geom_smooth` = `layer_smooth` = `layer_area` + `layer_path` + `transform_smooth`
* `geom_step` = `layer_path` + `transform_step`
* `geom_text` = `layer_text`
* `geom_tile` = `layer_rect` + `props(width ~ prop_band(), height ~ prob_band())`
* `geom_violin`
* `geom_vline` = `layer_path` + `props(y2 ~ prop_group("height"))`
