<!--
%\VignetteEngine{knitr}
%\VignetteIndexEntry{Marks}
-->

```{r, echo = FALSE}
library(knitr)
library(ggvis)
opts_chunk$set(comment = "#", error = TRUE, tidy = FALSE, eval = FALSE)
```

# Marks

The ggvis mark is responsible for the drawing of the data. Each type of mark 
effectively represents a different type of geometric object, and so in some
sense they are the equivalent of geoms in ggplot2. However, there are a far
fewer marks than geoms because marks tend to be more capable, and because they
never perform any statistical transformation.

`ggvis` exposes 
[five vega marks](https://github.com/trifacta/vega/wiki/Marks#color-references)
that have very similar ggplot2 equivalents:

* `mark_area` = `geom_ribbon`, `geom_area`
* `mark_line` = `geom_path`, `geom_line` 
* `mark_rect` = `geom_rect`, `geom_bar`, `geom_tile`
* `mark_symbol` = `geom_point`
* `mark_text` = `geom_text`

All marks start with `mark_` so that you can always use tab autocomplete to 
remind yourself of the options.

There are two marks that have no equivalent in ggplot2:

* `mark_arc` which allows you to draw arcs. You could accomplish something
  similar in ggplot2 with `geom_rect()` with `coord_polar()` but it was
  relatively inflexible and required contorting your brain a bit.
  
* `mark_image` allows you to place arbitrary images on a plot. There is 
  currently little additional support for this in ggvis, so you'll need to 
  make sure the images are publicly available somewhere. If you've found
  a compelling use-case for `mark_image`, let us know so we can consider 
  how to better support it.

There are two vega marks that are not currently implemented in ggvis:

* Support for the group mark will be added in future major release that will 
  add support for facetting/subplots)

* I can't currently see any use for the path mark, which takes svg path
  specifications as input.

## Creating a mark

All marks take two optional arguments which are only needed if you want to 
override the values inherited from parent:

* `props`, a `props()` object used to add any custom override for this layer
* `data`, an optional dataset to override the data inherited from the parent

For example, the following two plots are equivalent. In the first one, the 
data and props are defined at the top level and inherited by the mark, and in 
the second, they are defined in the mark itself. Note that the order of the 
argumetns to `ggvis()` and the mark function are different: this is because 
you'll usually set the data in top-level plot object, and usually only override
props in the mark.

```{r}
ggvis(mtcars, props(x ~ wt, y ~ mpg, stroke = "red"), 
  mark_symbol())

ggvis(mark_symbol(props(x ~ wt, y ~ mpg, stroke = "red"), mtcars))
```

In terms of the data hierarchy, marks are leaves: they can not have any 
children. See the data hierarchy vignette for more details.

## Mark details

The following sections describe particular details related to individual marks.

### `mark_rect`

If you have an ordinal scale, set width and/or height to `prop_band()` to 
occupy the complete band corresponding to that categorical value.

```{r}
data(diamonds, package = "ggplot2")
library(plyr)
counts <- count(diamonds, c("cut", "color"))

ggvis(counts, props(x ~ cut, y ~ colour, fill ~ freq),
  mark_rect(props(width ~ prop_band(), y ~ prop_band()))
```

### `mark_line`

`mark_line` produces lines if fill is empty, and polygons if it's set to a 
value.

```{r}
df <- data.frame(x = c(1, 1, 2, 2), y = c(2, 1, 1, 2))
ggvis(df, props(x ~ x, y ~ y), mark_line())
# Add a fill colour to make it a polygon
ggvis(df, props(x ~ x, y ~ y, fill = "red"), mark_line())
```

You can not currently set the component of lines to different colours: 
track progress at https://github.com/trifacta/vega/issues/122.

To create a line that spans the complete range of the plot, use `prop_group()`:

```{r}
ggvis(
  mark_symbol(props(x ~ disp, y ~ wt), mtcars),
  mark_line(props(y ~ mean(mtcars$wt), x = 0, x2 = prop_group()))
  mark_line(props(x ~ mean(mtcars$disp), y = 0, y2 = prop_group()))
)
```

Note that this won't work because `prop_group()` hasn't been written yet: track 
progress at https://github.com/rstudio/ggvis/issues/38.

### `mark_area`

You need to set two of `y`, `y2` and `height`:

```{r}
df <- data.frame(x = 1:10, y = (1:10) ^ 2)
ggvis(df, props(x ~ x, y ~ y, y2 ~ 0), mark_area())

# Set height in pixels
ggvis(df, props(x ~ x, y ~ y, height = 20), mark_area())
```

Height is only suitable for constant, because it does not have an obvious scale
associated with it. You could force height to use the `y` scale, but that 
doesn't work - the area hangs below the y line, and increasing the value of
height makes the area narrower! What's going on is that the underlying graphics
device has (0, 0) in the top-left corner, and so the y-scale is upside down. As
you increase height, it's mapped like a y variable so bigger values are further
away.

```{r}
ggvis(df, props(x ~ x, y ~ y, height = prop(80, scale = "y")), mark_area())
  
df <- data.frame(x = 1:10, y = (1:10) ^ 2)
ggvis(df, props(x ~ x, y ~ y), 
  mark_area(props(height = prop(input_slider(0, 100), scale = "y"))),
  mark_line(props(stroke = "red", strokeWidth = 10))
)
```

If you want to set the height in data units, you need to
compute the offsets yourself:

```{r}
ggvis(df, props(x ~ x, y ~ y - 2, y2 ~ y + 2), mark_area())
```

## Grouping

In ggplot2, grouping was a special aesthetic; in ggvis, grouping is a data 
transformation: use `by_group()` to split your data up into pieces given a 
specified variable:

```{r}
ggvis(mtcars, props(x ~ wt, y ~ mpg),
  mark_symbol(),
  mark_line(data = by_group("cyl"))
)
```

## Branches

In ggvis marks strictly represent drawing; a branch combines a mark with a
transformation, and often a set of default props. For example:

* `branch_histogram` = `mark_rect` + `transform_bin`
* `branch_freqpoly` = `mark_line` + `transform_bin`
* `branch_smooth` = (`mark_line` + `mark_area`) + `transform_bin`

All branches start with `branch_` and the source will typically be simple so is 
worth reading:

```{r, eval = TRUE}
branch_smooth
branch_histogram
```

If you find yourself using a set of marks commonly, it might be worth writing
your own branch function. Typically, the first argument should be `props` and
`...` should be passed on to the tranform function.

## Conversion from ggplot2

The following list provides a conversion from ggplot2 geoms to vega marks. 
However, because ggvis currently provides few transformations, many translations
don't currently exists - but they are on the roadmap and will be added over 
time.

* `geom_abline`
* `geom_area` = `mark_area` with `props(y ~ 0)`
* `geom_bar` = `mark_rect` with `props(y ~ 0, width = band())`
* `geom_bin2d` = `mark_rect` + `transform_bin2d`
* `geom_blank`
* `geom_boxplot`
* `geom_contour`
* `geom_crossbar`
* `geom_density` = `mark_line` + `transform_density`
* `geom_density2d`
* `geom_dotplot`
* `geom_errorbar` 
* `geom_errorbarh`
* `geom_freqpoly`
* `geom_hex`
* `geom_histogram` = `branch_histogram` = `mark_bar` + `transform_bin`
* `geom_hline` =  = `mark_line` + `props(x2 ~ proo_group("width"))`.
* `geom_jitter`
* `geom_line` = `mark_line` + `transform_sort`
* `geom_linerange`
* `geom_map`
* `geom_path` = `mark_line`
* `geom_point` = `mark_symbol`
* `geom_pointrange` 
* `geom_polygon` = `mark_line`
* `geom_quantile` = `mark_line` + `transform_quantreg`
* `geom_raster`
* `geom_rect` = `mark_rect`
* `geom_ribbon` = `mark_area`
* `geom_rug`
* `geom_segment`
* `geom_smooth` = `branch_smooth` = `mark_area` + `mark_line` + `transform_smooth`
* `geom_step` = `mark_line` + `transform_step`
* `geom_text` = `mark_text`
* `geom_tile` = `mark_rect` + `props(width ~ prop_band(), height ~ prob_band())`
* `geom_violin`
* `geom_vline` = `mark_line` + `props(y2 ~ prop_group("height"))`