---
title: "ggvis overview"
output: html_document
runtime: shiny
---

<!--
%\VignetteEngine{knitr}
%\VignetteIndexEntry{ggvis basics}
-->

```{r, echo = FALSE, message = FALSE}
library(knitr)
library(ggvis)
opts_chunk$set(comment = "#>", error = FALSE, tidy = FALSE)
opts_chunk$set(fig.width = 2.5, fig.height = 1.5, dpi = 100)
```

ggvis is a data visualization package for R which lets you:

* Declaratively describe data graphics with a syntax similar in spirit to 
  ggplot2.

* Create rich interactive graphics that you can play with locally in Rstudio
  or in your browser.

* Leverage [shiny](http://www.rstudio.com/shiny/)'s infrastructure to publish
  interactive graphics usable from any browser (either within your company
  or to the world).

The goal is to combine the best of R (e.g. every modelling function you can imagine) and the best of the web (everyone has a web browser). Data manipulation and transformation are done in R, and the graphics are rendered in a web browser, using [Vega](https://github.com/trifacta/vega/). For RStudio users, ggvis graphics display in a viewer panel, which is possible because RStudio is a web browser.

## Examples of ggvis graphics

Histogram:

```{r, echo = FALSE, fig.width = 4}
# Histogram
faithful %>% ggvis(~eruptions, fill := "#ffffdd", fill.hover := "#eebbbb") %>%
  layer_histograms(binwidth = 0.2) %>%
  add_axis("x", title = "eruptions") %>%
  add_axis("y", title = "count")
```

Scatterplot with smooth curve and interactive control:

```{r, echo = FALSE, fig.width = 3, fig.height = 3}
# Scatter plot with model fit line
mtcars %>%
  ggvis(~wt, ~mpg) %>%
  layer_points() %>%
  layer_smooths(span = input_slider(0.2, 1, 0.75, step = 0.05,
    label = "Smoothing span"))
```

## Getting started with ggvis

* See some [recipes](cookbook.html) for common tasks.
* Learn [ggvis basics](ggvis-basics.html).
* Read more about [interactive graphics](interactivity.html).

## More advanced topics

* How to control [axes and legends](axes-legends.html).
* How to use [properties and scales](properties-scales.html).
* Learn about [layers](layers.html).
* Plumbing: about ggvis's [data hierarchy](data-hierarchy.html).

## More about ggvis

* Differences and similarities to [ggplot2](ggplot2.html).
* The relationship between ggvis and [Vega](vega.html).

