---
title: "ggvis overview"
---

<!--
%\VignetteEngine{knitr}
%\VignetteIndexEntry{ggvis basics}
-->

```{r, echo = FALSE, message = FALSE}
library(knitr)
library(ggvis)
opts_chunk$set(comment = "#>", error = FALSE, tidy = FALSE)
opts_chunk$set(fig.width = 2.5, fig.height = 1.5, dpi = 100)
```

ggvis is a data visualization package for R which lets you:

* Declaratively describe data graphics with a syntax that is similar to ggplot2.
* View and interact with the graphics on your local computer, or use them in an interactive web application with [Shiny](http://www.rstudio.com/shiny/).

Data manipulation and transformation are done in R, and the graphics are rendered in a web browser, using [Vega](https://github.com/trifacta/vega/). For RStudio users, ggvis graphics display in a viewer panel, which is possible because RStudio is a web browser.


## Examples of ggvis graphics

Histogram:

```{r, echo = FALSE, message = FALSE, fig.width = 4}
# Histogram
faithful %>% ggvis(~eruptions, fill := "#ffffdd", fill.hover := "#eebbbb") %>%
  layer_histograms() %>%
  add_guide_axis("x", title = "eruptions") %>%
  add_guide_axis("y", title = "count")
```

Scatterplot with model fit line and interactive control:

```{r, echo = FALSE, message = FALSE}
# Scatter plot with model fit line
mtcars %>% 
  ggvis(~wt, ~mpg) %>% 
  layer_points() %>%
  layer_smooths()
```

## Getting started with ggvis

* See some [quick examples](quick-examples.html)
* Learn [ggvis basics](ggvis-basics.html)
* Read more about [interactive graphics](interactivity.html)

## More advanced topics

* How to control [axes and legends](axes-legends.html)
* How to use [properties and scales](properties-scales.html)
* Learn about [layers](layers.html)
* Plumbing: about ggvis's [data hierarchy](data-hierarchy.html)

## More about ggvis

* Differences and similarities to [ggplot2](ggplot2.html)
* The relationship between ggvis and [Vega](vega.html)

