---
title: Properties and scales
output:
  ggvis:::html_document:
    toc: true
---

<!--
%\VignetteEngine{knitr}
%\VignetteIndexEntry{Properties and scales}
-->

```{r, echo = FALSE, message = FALSE}
library(knitr)
library(ggvis)
opts_chunk$set(comment = "#>", error = FALSE, tidy = FALSE)
opts_chunk$set(fig.width = 3.5, fig.height = 2.5, dpi = 100)
```

Understanding how properties and scales work in ggvis an important step to
mastering basic static plots, and is also necessary for great interactive
graphics.

In this chapter, you will learn:

* the convenient `props()` wrapper which makes the most common types of
  property mappings available through a concise interface

* the `prop()` function, which is more verbose, but gives you full control
  over all options

* what scales do, how properties and scales are connected, how you can
  override the defaults.

Note that unlike ggplot2, scales do not control the appearance of their guides:
see the [axes and legends](axes-legends.html) vignette for how to customise their display.

## The `props()` wrapper

Every ggvis mark is associated with a set of properties that governs how it is displayed. These properties can be constant values (like 5, "blue", or "square"), or mapped to variables in your dataset. ggplot2 syntax made a distinction between mapping variables and setting constants. For example, in ggplot2, you might say:

```{r eval = FALSE}
geom_point(aes(x = wt, y = mpg), colour = "red", size = 5)
```

But in ggvis, everything is a property:

```{r}
layer_path(props(x = ~wt, y = ~ mpg, stroke := "red", strokeWidth := 5))
```

This section introduces `props()`, a convenient function for creating property objects and binding them to property names. The next chapter shows you how to create the property objects by hand, giving you more control over the specification at the cost of some extra typing.

### Setting vs mapping

Props distinguishes between setting values to constants and mapping values to a variable based on whether you use `=` (mapping) or `:=` (setting). The syntax is designed to remind you that, like a linear model, the properties specification isn't using the value of the variable `wt`, but its name.

* `props(x = ~displ, y = ~mpg)`: map x to engine displacement and y to miles
  per gallon

* `props(stroke := "red", fill := "blue")`: set the stroke colour to red and the
  fill colour to blue.

* `props(x = ~displ, y := 10)`: map the x-position to displacement and set
  the y position to 10 pixels.

Those examples follow a common pattern: `=` is always scaled, and `:=` is never scaled. `~` is always used with the name of the variable. What if you try the opposite?

* `props(x = 0)`: sets the x position to the data value 0

It's also possible provide a scaled constant instead of a raw constant.
That's useful when you want to label different layers in a plot:

```{r, eval = FALSE}
ggvis(mtcars, props(x = ~wt, y = ~mpg)) +
  layer_point() +
  layer_smooth(props(stroke = "lm"), method = "lm") +
  layer_smooth(props(stroke = "loess"))
```

(Note: this isn't currently supported in ggvis because of a limitation
of vega. See https://github.com/rstudio/ggvis/issues/29 for progress.)

### Valid properties

Not all marks support all properties. The complete list of all properties is
available in `?marks`, and mark functions will check that you've supplied them
valid properties, and will endeavour to provide you a helpful suggestion:

```{r, eval = FALSE}
layer_line(props(strke = ~cyl))
# Error: Unknown properties: strke.
# Did you mean: stroke?
```

Note that properties supplied to `ggvis` or `layer` can't be checked
automatically on creation, because they may be modified by other nodes along
the pipeline (or the node structure of the graph, including the marks, might
be changed after plot creation). These will only be checked once the plot
is rendered.

```{r, eval = FALSE}
# No error on creation
p <- ggvis(mtcars, props(x = ~disp, y = ~mpg, strke = ~cyl), layer_point())
# Error when attempt to display it
print(p)
# Error message currently not informative: see
# https://github.com/rstudio/ggvis/issues/31 for progress
```

### Capture of local variables

Variable properties can refer to both variables in the dataset and variables in
the local environment:

```{r, results = 'asis'}
df <- data.frame(x = 1:10)
f <- function(n) {
  ggvis(df, props(x = ~x, y = ~x ^ n), layer_path())
}
f(1)
f(2)
```

Technically, ggvis uses the environment captured by the formula when it is
created, which may be important if you're generating formulas in one function
and using them in another. You can always override the environment by calling
`prop()` and supplying the `env` argument.

## `prop()`

A prop has two key properties:

* value: which can be a constant, the name of a variable (or an expression
  involving one or more variables), or an interactive input. An interactive
  input must yield either a constant, or an variable name/expression.

* the scale: if scaled, a vega scale is in charge of converting the data
  value to something that makes sense as a visual property. If unscaled, the
  value is used as is.

  Unscaled is the equivalent of using `scale_identity` in ggplot2.

Type        | Scaled | Common use | In props
------------|--------|-------------------------
variable    | yes    | Map variable to property | `x = ~mpg`
variable    | no     | Variable already contains visual values | `x := ~colour`
constant    | yes    | Annotation position in data space | `x = 1`
constant    | no     | Pixel location, or exact colour | `x := 1`

### Special evaluation and variables

`prop()` doesn't do any special evaluation which means that you if you want a
variable, you need to supply a quoted expression or a one-sided formula:

```{r}
prop(quote(x))
prop(~ x)
```

If you have the name of a variable as a string, you can convert it a name with
`as.name()`:

```{r}
var <- "mpg"
prop(as.name(var))
```

If you have an R expression as a string, `parse()` it then extract the first
element of the result:

```{r}
expr <- "mpg / wt"
prop(parse(text = expr)[[1]])
```

## Properties -> scales

Like in ggplot2, scales control the mapping between data values and values
interpreted by the drawing device. Scales are added to the top-level plot
specification and are usually created with `dscale` (short for default scale):

```{r, results = 'asis'}
# Override the default data limits:
ggvis(mtcars, props(x = ~disp, y = ~wt)) +
  layer_point() +
  dscale("x", "numeric", domain = c(50, 500), nice = FALSE) +
  dscale("y", "numeric", domain = c(0, 6), nice = FALSE)
```

Compared to ggplot2, ggvis has far fewer scales (3 vs 70), with each function
doing much more. The three basic scales in ggvis, corresponding to the
[three basic vega scales][vega-scales] are:

* `scale_quantitative`: for quantitative, numeric values
* `scale_ordinal`: for qualitative, categorical values
* `scale_time`: for date/time values

The vega scales are in turn relatively simple wrappers for
[D3 scales][d3-scales] so if you can't find the details in these docs, or in
the vega docs, you may need to look in the D3 docs. Fortunately the arguments
are by and large named the same across all three systems, although ggvis uses
underscores while vega and D3 use camelCase.

Each (scaled) property needs a scale. By default, scales are added so that
every scaled property gets a scale with the same name as the property (with a
few exceptions `y2` to `y`, `x2` to `x`, `fillOpacity` and `strokeOpacity` to
`opacity` and so on.). See `?add_default_scales` for details.

### Scale arguments

The scales share the following arguments:

* `name`: a string identifier for the scale. By default, this is the name of the
  property it is associated with - i.e. the scale for the x values is called
  "x", but it doesn't have to be, and the examples below show some cases where
  you need this extra flexibility.

* `domain`: the input data values to the scale. If left blank, these will be
  learned from the properties that use this scale. But you can override it
  if you want to expand or restrict, or you want to match domains across
  multiple plots so that they're easier to compare. `domain` is equivalent to
  `limits` in R graphics. For quantiative scales, you can use a missing value
  to only constrain one side of the domain.

* `range`: the output visual values from the scale. For quantiative scales this
  is usually a vector of length two, for ordinal scales, it's a vector the same
  length as the domain. Vega interprets some special values: "width", "height",
  "shapes", "category10" (10 categorical colours).

* `reverse`: a boolean flag that will flip the order of the range.

`scale_quantitative` and `scale_time` also share a few other properties:

* `nice`: use nice axis labels? The algorithm is described in the
  [D3 document][d3-nice]

* `zero`: include a zero in the scale?

* `clamp`: clamp any values outside the domain to the min/max?

### Default scales

Because the range of a scale is usually determined by the type of variable,
ggvis provides the `dscale` function to automatically retrieve a default scale
given a property name and variable type:

```{r}
dscale("x", "numeric")
dscale("y", "numeric")
dscale("shape", "nominal")
```

You can also provide other arguments to the underlying scale:

```{r}
dscale("x", "numeric", domain = c(10, 100))
dscale("x", "numeric", trans = "log")
```

So `dscale()` is usually a better way to create new scales that starting from
the underlying scale objects.

### Custom scales

You can add your own scales for properties that don't otherwise have defaults.
For example, imagine you wanted to use the font of a label to represent some
data. There's no default scale for font, but you could create one yourself:

```{r, results = 'asis'}
df <- data.frame(x = runif(5), y = runif(5),
  labels = c("a", "b", "b", "a", "b"))
ggvis(df, props(x = ~x, y = ~y, text := ~labels, font = ~labels,
    fontSize := 40)) +
  layer_text() +
  scale_ordinal("font", range = c("Helvetica Neue", "Times New Roman"))
```

Note the use of `text := ~labels`: we don't want to scale the labels - the raw values already make sense in the visual space.

### Multiple scales for one property

Generally, you will override the default name of a scale in order to use
more scales than the default. You could do this in order to create a dual-axis
chart (which is generally a bad idea - read [this paper][dual-axis] for more details).
If you do this, you will also need to add a scale object.

```{r, results = 'asis'}
ggvis(mtcars, props(y = ~mpg)) +
  layer_point(props(x = prop(~disp, scale = "xdisp"))) +
  layer_point(props(x = prop(~wt, scale = "xwt"), fill := "blue")) +
  dscale("x", "numeric", name = "xdisp") +
  dscale("x", "numeric", name = "xwt") +
  guide_axis("x", "xdisp", orient = "bottom") +
  guide_axis("x", "xwt", orient = "bottom", offset = 20, properties =
    list(labels = props(fill := "blue")))
```

### Multiple properties for one scale

You could also force ggplot2 to use the same scale for properties that would
otherwise use different scales. I'm not sure of a useful example of this,
except to force `stroke` and `fill` to use the same scale:

```{r, results = 'asis'}
df <- data.frame(x = 1:5, y = 1:5, a = runif(5), b = -runif(5))
p <- props(x = ~x, y = ~y, stroke = ~a, fill = ~b, strokeWidth := 5, size := 1000)
ggvis(df, p, layer_point(),
  guide_legend(fill = "stroke", properties = list(legend = props(y := 50)))
)

p2 <- props(x = ~x, y = ~y, stroke = ~a, fill = prop(~b, scale = "stroke"),
  strokeWidth := 5, size := 1000)
ggvis(df, p2, layer_point())
```

In this case we don't need to manually add the correct scale, because ggvis
has detected it for us automatically.

### Property values

Vega renders either svg or canvas, but fortunately most properties are shared
across svg or canvas. The following list describes what the property values
mean and the set of allowable values.

* `x`, `x2`, `width`, `y`, `y2`, `height`, `strokeWidth`, `innerRadius`,
  `outerRadius`: pixels.  Note that by convention (0, 0) is located in the
  top-left, so y values are relative to the top of the screen and x values
  are relative to the left of the screen (as opposed to R where (0,0) is on
  the bottom right). Pixel positions should be greater than 0.

* `size`: area, in pixels. Greater than 0.

* `opacity`, `fillOpacity`, `strokeOpacity`: a number between 0 and 1

* `stroke`, `fill`: colours

* `startAngle`, `endAngle`: angle in radians

* `interpolate`: "linear", "step-before", "step-after", "basis", "basis-open",
  "cardinal", "cardinal-open", "monotone".  See the [D3 docs][interpolate-docs]
  for what they mean.

* `tension`: a number between 0 and 1 that controls a tension parameter to
  some interpolants.See the [D3 docs][tension-docs] for more details.

* `url`: a url.

* `align`: "left", "right", "center".

* `baseline`: "top", "middle", "bottom"

* `text`: a string

* `dx`, `dy`: pixel offsets from anchor point

* `angle`: an angle in degrees

* `font`: the name of a font available from the browser.

* `fontWeight`: a font weight, as a string ("normal", "bold", "lighter",
  "bolder") or number (100, 200, 300, 400, 500, 600, 700, 800, 900).

* `fontStyle`: "normal", "italic", "oblique"


[dual-axis]: https://www.lri.fr/~isenberg/publications/papers/Isenberg_2011_ASO.pdf
[interpolate-docs]: https://github.com/mbostock/d3/wiki/SVG-Shapes#wiki-line_interpolate
[tension-docs]: https://github.com/mbostock/d3/wiki/SVG-Shapes#wiki-line_tension
[d3-nice]: https://github.com/mbostock/d3/wiki/Quantitative-Scales#wiki-linear_nice
[vega-scales]: https://github.com/trifacta/vega/wiki/Scales
[d3-scales]: https://github.com/mbostock/d3/wiki/Scales
