<!--
%\VignetteEngine{knitr}
%\VignetteIndexEntry{Properties and scales}
-->

```{r, echo = FALSE}
library(knitr)
library(ggvis)
opts_chunk$set(comment = "#", error = TRUE, tidy = FALSE, eval = FALSE)
```

# Properties and scales

Understanding how properties and scales work in ggvis an important step to 
mastering basic static plots, and is also necessary for great interactive
graphics.

In this chapter, you will learn:

* the convenient `props()` wrapper which makes the most common types of 
  property mappings available through a concise interface
  
* the `prop()` function, which is more verbose, but gives you full control
  over all options
  
* how properties are associated with scales, and how you can override 
  those choices

Note that unlike ggplot2, scales do not control the appearance of their guides:
see the axes and legends vignette for how to customise the display of those 
components.

## The `props()` wrapper

### Getting started

`props()` has a similar purpose to ggplot2's `aes()` - it associates variables
in your datasets, with visual properties in the graphic. However, `props()` 
has been design a little differently to `aes()`, because as well as including
all the mappings between variables and properties, it also includes fixed 
values. For example, in ggplot2, you might say:

```R
geom_point(aes(x = wt, y = mpg), colour = "red", size = 5)
```

but in ggvis, you say:

```R
mark_line(props(x ~ wt, y ~ mpg, stroke = "red", size = 5))
```

To distinguish between variables that you are _mapping_, and values that you
are _setting_, you use `~` or `=` respectively. The `~` is an indication that
props isn't using the value of the variable `wt`, but it's name. 

(This obviously requires a little non-standard evaluation - if you want to 
create mappings given the names of variables as a string (e.g `x <- "mpg")
you'll need to use the more explicit `prop()` function described below.)

It's also possible to "map" a constant, which effectively adds a new column to
the data containing a constant value, and then maps the variable to that. 
That's useful when you want to label different layers in a plot:

```R
ggvis(mtcars, props(x ~ wt, y ~ mpg), 
  mark_symbol(),
  branch_smooth(props(stroke = "lm"), method = lm),
  branch_smooth(props(stroke = "loess"))
)
```

(Note this isn't currently supported in ggvis because of a limitation 
of vega. See https://github.com/rstudio/ggvis/issues/29 for progress.)

Trying to set a property to a variable won't work `props(x = wt)`, unless there
is a variable called `wt` defined in the local environment, in which case its
value will be used as constant.


### Valid properties

Not all marks support all properties. The complete list of all properties is 
available in `?marks`, and mark functions will check that you've supplied them
valid properties, and will endeavour to provide you a helpful suggestion:

```{r, eval = TRUE}
mark_line(props(strke ~ cyl))
```

Note that properties supplied to `ggvis` or `node` can't be checked 
automatically on creation, because they may be modified by other nodes along
the pipeline (or the node structure of the graph, including the marks, might
be changed after plot creation). These will only be checked once the plot
is rendered.

```{r}
# No error on creation
p <- ggvis(mtcars, props(x ~ disp, y ~ mpg, strke ~ cyl), mark_symbol())
# Error when attempt to display it
print(p)
# Error message currently not informative: see 
# https://github.com/rstudio/ggvis/issues/31 for progress
```

### Capture of local variables

(Currently doesn't work)

## `prop()`

A prop has three key properties:

* constant/variable: if constant, the value is used as is. If variable, it is
  evaluated in the context of the data frame. 

* scaled/unscaled: if scaled, a vega scale is in charge of converting the data
  value to something that makes sense as a visual property. If unscaled, the
  value is used as is. Unscaled is the equivalent of using `scale_identity` in
  ggplot2. Scaled defaults to 
  
* reactive/static: automatically detected from the input. If it's a delayed
  reactive, it will be re-evaluted multiple times during the life-time of the
  plot, as the underlying input changes.

This gives rise to eight possible combinations:

Reactive | Variable | Scaled | Common use | In props
---------|----------|--------|-------------------------
FALSE    | FALSE    | FALSE  | Pixel location, or exact colour | `x = 1`
FALSE    | FALSE    | TRUE   | Annotation position in data space | `x ~ 1`
FALSE    | TRUE     | FALSE  | Variable already contains visual values | 
FALSE    | TRUE     | TRUE   | Map variable to property | `x ~ mpg`
TRUE     | FALSE    | FALSE  | Choose pixel location or color | `x = input_slider()`
TRUE     | FALSE    | TRUE   | Choose position in data space |
TRUE     | TRUE     | FALSE  | Choose visual variable to display
TRUE     | TRUE     | TRUE   | Choose variable to display |

`props()` uses the heuristics to avoid you having to explicit set these values
for the most common uses. It uses the following heuristics to set:

* `~`: `scaled = TRUE`, `reactive = FALSE`. `constant = TRUE` if the RHS of the 
  formula is an atomic vector, `FALSE` if LHS is a call or name. 

* `=`: `scaled = FALSE`. `reactive = TRUE` if input is a delayed reactive,
  otherwise `FALSE`, `scaled = !reactive`

If you need to do something different you can explicitly create the `prop()` 
object.  For example, if you want to make a non-constant scaled reactive, you
could do:

```R
vars <- input_select(c("disp", "wt"))
gigvis(mtcars, props(x = prop(vars, constant = FALSE), y ~ mpg), 
  mark_symbol())
```

### Special evaluation and variables

`prop()` doesn't do any special evaluation which means that you if you want a
non-constant scale, you need to supply a quoted expression:

```R
prop(quote(x))
var <- quote(x)
prop(var)
```

If you have the name of a variable as a string, you can convert it a name with
`as.name()`:

```R
var <- "mpg"
prop(as.name(var))
```

If you have an R expression as a string, `parse()` it then extract the first
element of the result:

```R
expr <- "mpg / wt"
prop(parse(text = expr)[[1]])
```

## Properties -> scales

Like in ggplot2, scales control the mapping between data values and values 
interpreted by the drawing device. Scales are added to the top-level plot 
specification and are usually created with `dscale` (short for default scale):

```R
# Override the default data limits:
ggvis(mtcars, props(x ~ disp, y ~ wt), 
  mark_symbol(),
  dscale("x", "numeric", domain = c(50, 500), nice = FALSE),
  dscale("y", "numeric", domain = c(0, 6), nice = FALSE)
)
```

Compared to ggplot2, ggvis has far fewer scales (3 vs 70), with each function 
doing much more. The three basic scales in ggvis, corresponding to the 
[three basic vega scales][vega-scales] are:

* `scale_quantitative`: for quantitative, numeric values
* `scale_ordinal`: for qualitative, categorical values
* `scale_time`: for date/time values

The vega scales are in turn relatively simple wrappers for 
[D3 scales][d3-scales] so if you can't find the details in these docs, or in 
the vega docs, you may need to look in the D3 docs. Fortunately the arguments 
are by and large named the same across all three systems, although ggvis uses
underscores while vega and D3 use camelCase.

Each (scaled) property needs a scale. By default, scales are added so that
every scaled property gets a scale with the same name as the property (with a 
few exceptions `y2` to `y`, `x2` to `x`, `fillOpacity` and `strokeOpacity` to 
`opacity` and so on.). See `?add_default_scales` for details.

### Scale arguments

The scales share the following arguments:

* `name`: a string identifier for the scale. By default, this is the name of the 
  property it is associated with - i.e. the scale for the x values is called 
  "x", but it doesn't have to be, and the examples below show some cases where
  you need this extra flexibility.
  
* `domain`: the input data values to the scale. If left blank, these will be
  learned from the properties that use this scale. But you can override it 
  if you want to expand or restrict, or you want to match domains across 
  multiple plots so that they're easier to compare. `domain` is equivalent to
  `limits` in R graphics. For quantiative scales, you can use a missing value
  to only constrain one side of the domain.
  
* `range`: the output visual values from the scale. For quantiative scales this
  is usually a vector of length two, for ordinal scales, it's a vector the same
  length as the domain. Vega interprets some special values: "width", "height",
  "shapes", "category10" (10 categorical colours).

* `reverse`: a boolean flag that will flip the order of the range.

`scale_quantitative` and `scale_time` also share a few other properties:

* `nice`: use nice axis labels? The algorithm is described in the 
  [D3 document][d3-nice]

* `zero`: include a zero in the scale? 

* `clamp`: clamp any values outside the domain to the min/max?

### Default scales

Because the range of a scale is usually determined by the type of variable, 
ggvis provides the `dscale` function to automatically retrieve a default scale
given a property name and variable type:

```{r, eval = TRUE}
dscale("x", "numeric")
dscale("y", "numeric")
dscale("shape", "nominal")
```

You can also provide other arguments to the underlying scale:

```{r, eval = TRUE}
dscale("x", "numeric", domain = c(10, 100))
dscale("x", "numeric", trans = "log10")
```

So `dscale()` is usually a better way to create new scales that starting from
the underlying scale objects.

### Custom scales

You can add your own scales for properties that don't otherwise have defaults.
For example, imagine you wanted to use the font of a label to represent some
data. There's no default scale for font, but you could create one yourself:

```{r}
df <- data.frame(x = runif(5), y = runif(5), 
  labels = c("a", "b", "b", "a", "b"))
ggvis(df, 
  props(x ~ x, y ~ y, text = prop(quote(labels), scale = F), font ~ labels, 
    fontSize = 40), 
  mark_text(),
  scale_ordinal("font", range = c("Helvetica Neue", "Times New Roman")))
```

### Multiple scales for one property

Generally, you will override the default name of a scale in order to use 
more scales than the default. You could do this in order to create a dual-axis
chart (which is a bad idea - read [this paper][dual-axis] for more details).
If you do this, you will also need to add a scale object.

```R
ggvis(mtcars, props(y ~ mpg), 
  mark_symbol(props(x = prop(quote(disp), scale = "xdisp"))),
  mark_symbol(props(x = prop(quote(wt), scale = "xwt"), stroke = "blue")),
  dscale("x", "numeric", name = "xdisp"),
  dscale("x", "numeric", name = "xwt"),
  guide_axis("x", "xdisp", orient = "bottom"),
  guide_axis("x", "xwt", orient = "bottom", properties = list(labels = props(fill = "blue")))
)
```

### Multiple properties for one scale

You could also force ggplot2 to use the same scale for properties that would
otherwise use different scales. I'm not sure of a useful example of this, 
except to force `stroke` and `fill` to use the same scale:

```R
df <- data.frame(x = 1:5, y = 1:5, a = runif(5), b = -runif(5))
p <- props(x ~ x, y ~ y, stroke ~ a, fill ~ b, strokeWidth = 5, size = 1000)
ggvis(df, p, mark_symbol(),
  guide_legend(fill = "stroke", properties = list(legend = props(y = 50)))
)

p2 <- props(x ~ x, y ~ y, stroke ~ a, fill = prop(quote(b), scale = "stroke"), 
  strokeWidth = 5, size = 1000)
ggvis(df, p2, mark_symbol())
```

In this case we don't need to manually add the correct scale, because ggvis
has detected it for us automatically.


### Property values

Vega renders either svg or canvas, but fortunately most properties are shared 
across svg or canvas. The following list describes what the property values 
mean and the set of allowable values.

* `x`, `x2`, `width`, `y`, `y2`, `height`, `strokeWidth`, `innerRadius`, 
  `outerRadius`: pixels.  Note that by convention (0, 0) is located in the 
  top-left, so y values are relative to the top of the screen and x values
  are relative to the left of the screen (as opposed to R where (0,0) is on 
  the bottom right). Pixel positions should be greater than 0.

* `size`: area, in pixels. Greater than 0.

* `opacity`, `fillOpacity`, `strokeOpacity`: a number between 0 and 1

* `stroke`, `fill`: colours

* `startAngle`, `endAngle`: angle in radians

* `interpolate`: "linear", "step-before", "step-after", "basis", "basis-open", 
  "cardinal", "cardinal-open", "monotone".  See the [D3 docs][interpolate-docs]
  for what they mean. 
  
* `tension`: a number between 0 and 1 that controls a tension parameter to 
  some interpolants.See the [D3 docs][tension-docs] for more details.
  
* `url`: a url.

* `align`: "left", "right", "center".

* `baseline`: "top", "middle", "bottom"

* `text`: a string

* `dx`, `dy`: pixel offsets from anchor point

* `angle`: an angle in degrees

* `font`: the name of a font available from the browser. 

* `fontWeight`: a font weight, as a string ("normal", "bold", "lighter", 
  "bolder") or number (100, 200, 300, 400, 500, 600, 700, 800, 900).

* `fontStyle`: "normal", "italic", "oblique"


[dual-axis]: https://www.lri.fr/~isenberg/publications/papers/Isenberg_2011_ASO.pdf
[interpolate-docs]: https://github.com/mbostock/d3/wiki/SVG-Shapes#wiki-line_interpolate
[tension-docs]: https://github.com/mbostock/d3/wiki/SVG-Shapes#wiki-line_tension
[d3-nice]: https://github.com/mbostock/d3/wiki/Quantitative-Scales#wiki-linear_nice
[vega-scales]: https://github.com/trifacta/vega/wiki/Scales
[d3-scales]: https://github.com/mbostock/d3/wiki/Scales
