<!--
%\VignetteEngine{knitr}
%\VignetteIndexEntry{Properties and scales}
-->

# Properties and scales

Understanding how properties and scales work in ggvis an important step to 
mastering basic static plots, and is also necessary for great interactive
graphics.

In this chapter, you will learn:

* the convenient `props()` wrapper which makes the most common types of 
  property mappings available through a concise interface
  
* the `prop()` function, which is more verbose, but gives you full control
  over all options
  
* how properties are associated with scales, and how you can override 
  those choices
  
## The `props()` wrapper

### Getting started

`props()` has a similar purpose to ggplot2's `aes()` - it associates variables
in your datasets, with visual properties in the graphic. However, `props()` 
has been design a little differently to `aes()`, because as well as including
all the mappings between variables and properties, it also includes fixed 
values. For example, in ggplot2, you might say:

```R
geom_point(aes(x = wt, y = mpg), colour = "red", size = 5)
```

but in ggvis, you say:

```R
mark_line(props(x ~ wt, y ~ mpg, stroke = "red", size = 5))
```

To distinguish between variables that you are _mapping_, and values that you
are _setting_, you use `~` or `=` respectively. The `~` is an indication that
props isn't using the value of the variable `wt`, but it's name. 

(This obviously requires a little non-standard evaluation - if you want to 
create mappings given the names of variables as a string (e.g `x <- "mpg")
you'll need to use the more explicit `prop()` function described below.)

It's also possible to "map" a constant, which effectively adds a new column to
the data containing a constant value, and then maps the variable to that. 
That's useful when you want to label different layers in a plot:

```R
ggvis(mtcars, props(x ~ wt, y ~ mpg), 
  mark_symbol(),
  branch_smooth(props(stroke = "lm"), method = lm),
  branch_smooth(props(stroke = "loess"))
)
```

(Note this isn't currently supported in ggvis because of a limitation 
of vega. See https://github.com/rstudio/ggvis/issues/29 for progress.)

Trying to set a property to a variable won't work `props(x = wt)`, unless there
is a variable called `wt` defined in the local environment, in which case its
value will be used as constant.


### Valid properties

Not all marks support all properties. The complete list of all properties is 
available in `?marks`, and mark functions will check that you've supplied them
valid properties, and will endeavour to provide you a helpful suggestion:

```{r, eval = TRUE}
mark_line(props(strke ~ cyl))
```

Note that properties supplied to `ggvis` or `node` can't be checked 
automatically on creation, because they may be modified by other nodes along
the pipeline (or the node structure of the graph, including the marks, might
be changed after plot creation). These will only be checked once the plot
is rendered.

```{r, eval = FALSE}
# No error on creation
p <- ggvis(mtcars, props(x ~ disp, y ~ mpg, strke ~ cyl), mark_symbol())
# Error when attempt to display it
print(p)
# Error message currently not informative: see 
# https://github.com/rstudio/ggvis/issues/31 for progress
```

### Capture of local variables

(Currently doesn't work)

## `prop()`

A prop has three key properties:

* constant/variable: if constant, the value is used as is. If variable, it is
  evaluated in the context of the data frame. 

* scaled/unscaled: if scaled, a vega scale is in charge of converting the data
  value to something that makes sense as a visual property. If unscaled, the
  value is used as is. Unscaled is the equivalent of using `scale_identity` in
  ggplot2. Scaled defaults to 
  
* reactive/static: automatically detected from the input. If it's a delayed
  reactive, it will be re-evaluted multiple times during the life-time of the
  plot, as the underlying input changes.

This gives rise to eight possible combinations:

Reactive | Variable | Scaled | Common use | In props
---------|----------|--------|-------------------------
FALSE    | FALSE    | FALSE  | Pixel location, or exact colour | `x = 1`
FALSE    | FALSE    | TRUE   | Annotation position in data space | `x ~ 1`
FALSE    | TRUE     | FALSE  | Variable already contains visual values | 
FALSE    | TRUE     | TRUE   | Map variable to property | `x ~ mpg`
TRUE     | FALSE    | FALSE  | Choose pixel location or color | `x = input_slider()`
TRUE     | FALSE    | TRUE   | Choose position in data space |
TRUE     | TRUE     | FALSE  | Choose visual variable to display
TRUE     | TRUE     | TRUE   | Choose variable to display |

`props()` uses the heuristics to avoid you having to explicit set these values
for the most common uses. It uses the following heuristics to set:

* `~`: `scaled = TRUE`, `reactive = FALSE`. `constant = TRUE` if the RHS of the 
  formula is an atomic vector, `FALSE` if LHS is a call or name. 

* `=`: `scaled = FALSE`. `reactive = TRUE` if input is a delayed reactive,
  otherwise `FALSE`, `scaled = !reactive`

If you need to do something different you can explicitly create the `prop()` 
object.  For example, if you want to make a non-constant scaled reactive, you
could do:

```R
vars <- input_select(c("disp", "wt"))
gigvis(mtcars, props(x = prop(vars, constant = FALSE), y ~ mpg), 
  mark_symbol())
```

### Special evaluation and variables

`prop()` doesn't do any special evaluation which means that you if you want a
non-constant scale, you need to supply a quoted expression:

```R
prop(quote(x))
var <- quote(x)
prop(var)
```

If you have the name of a variable as a string, you can convert it a name with
`as.name()`:

```R
var <- "mpg"
prop(as.name(var))
```

If you have an R expression as a string, `parse()` it then extract the first
element of the result:

```R
expr <- "mpg / wt"
prop(parse(text = expr)[[1]])
```

## Properties -> scales

Each (scaled) property is associated with a scale. The default scale associated
with a property has the same name as the property (with a few exceptions `y2` 
to `y`, `x2` to `x`, `fillOpacity` and `strokeOpacity` to `opacity` and so on.)

Generally, you will override the default name of a scale in order to use 
more scales than the default. You could do this in order to create a dual-axis
chart (which is a bad idea - read [this paper][dual-axis] for more details).
If you do this, you will also need to add a scale object.

```R
ggvis(mtcars, props(y ~ mpg), 
  mark_symbol(props(x = prop(quote(disp), scale = "xdisp"))),
  mark_symbol(props(x = prop(quote(wt), scale = "xwt"), stroke = "blue")),
  dscale("x", "numeric", name = "xdisp"),
  dscale("x", "numeric", name = "xwt"),
  axis("x", "xdisp", orient = "bottom"),
  axis("x", "xwt", orient = "bottom", properties = list(labels = props(fill = "blue")))
)
```

You could also force ggplot2 to use the same scale for properties that would
otherwise use different scales. I'm not sure of a useful example of this, 
except to force `stroke` and `fill` to use the same scale:

```R
df <- data.frame(x = 1:5, y = 1:5, a = runif(5), b = -runif(5))
p <- props(x ~ x, y ~ y, stroke ~ a, fill ~ b, strokeWidth = 5, size = 1000)
ggvis(df, p, mark_symbol(),
  legend(fill = "stroke", properties = list(legend = props(y = 50)))
)

p2 <- props(x ~ x, y ~ y, stroke ~ a, fill = prop(quote(b), scale = "stroke"), 
  strokeWidth = 5, size = 1000)
ggvis(df, p2, mark_symbol())
```

In this case we don't need to manually add the correct scale, because ggvis
has detected it for us automatically.

[dual-axis]: https://www.lri.fr/~isenberg/publications/papers/Isenberg_2011_ASO.pdf