---
title: "Quick ggvis examples"
---

<!--
%\VignetteEngine{knitr}
%\VignetteIndexEntry{ggvis basics}
-->

```{r, echo = FALSE, message = FALSE}
library(knitr)
opts_chunk$set(comment = "#>", error = FALSE, tidy = FALSE)
opts_chunk$set(fig.width = 2, fig.height = 1.25, dpi = 100)
```

## qvis and ggvis

There are two primary functions in ggvis that are used to create plots: `qvis()` and `ggvis()`.  `qvis()` provides a streamlined interface which is suitable for simple plots; when you need more control over plots, `ggvis()` may be more appropriate.

In some of the simpler examples below, we'll show equivalent code for `qvis()` and `ggvis()`.

# Basic plots

First, load ggvis and dplyr:

```{r, message = FALSE}
library(ggvis)
library(dplyr)
```

## Scatterplots

We'll use the built-in `mtcars` data set, and look at two columns of interest, `mpg`, and `wt`:

```{r}
# The first few rows of mtcars
head(mtcars)
```

A basic scatter plot:

```{r, results = 'asis', message = FALSE}
mtcars %>% qvis(~wt, ~mpg)
mtcars %>% ggvis(~wt, ~mpg) %>% layer_points()
```

Smaller points, a different shape, a different outline (stroke) color, and empty fill:

```{r, results = 'asis', message = FALSE}
mtcars %>% 
  qvis(~wt, ~mpg, size := 25, shape := "diamond", stroke := "red", fill := NA)
mtcars %>% 
  ggvis(~wt, ~mpg) %>% 
  layer_points(size := 25, shape := "diamond", stroke := "red", fill := NA)
```


## Bar graphs

We'll use the built-in `pressure` data set for these examples:

```{r}
# The first few rows
head(pressure)
```

When the variable on the x axis is continuous (e.g., numeric or date-time):

```{r, results = 'asis', message = FALSE, fig.width = 4}
# Note that y2 is the bottom of the bar, and the width is specified in pixels
pressure %>% 
  ggvis(~temperature, ~pressure, y2 = 0) %>% 
  layer_rects(width := 10)
```

In the previous example, the width of each bar was specified in pixels. If you change the width of the plot, you'll see that the bars remain 10 pixels wide.

Instead of setting the width in pixels, it's possible to set the width in the scaled data space, by specifying the `x` and `x2` values for each bar.In this example, each bar is 20 wide in the x scale, so the bars touch:

```{r, results = 'asis', message = FALSE, fig.width = 4}
pressure %>% 
  ggvis(~temperature - 10, x2 = ~temperature + 10, ~pressure, y2 = 0) %>% 
  layer_rects()
```

When the variable on the x axis is categorical (e.g., factor or character):

```{r, results = 'asis', message = FALSE, fig.width = 4}
# First, modify the pressure data set so that the x variable is a factor
pressure2 <- pressure %>% mutate(temperature = factor(temperature))

pressure2 %>% ggvis(~temperature, ~pressure) %>%
  layer_rects(y2 = 0, width = band()) %>%
  set_dscale("x", "nominal", padding = 0, points = FALSE)
```

(Notice that in this example, the x values aren't sorted quite right: they're sorted lexically, by the first, second, and then third digit, instead of by the numeric value. This is due to a bug in ggvis.)

## Line graphs

```{r, results = 'asis', message = FALSE}
pressure %>% ggvis(~temperature, ~pressure) %>% layer_lines()
```

Lines with points:

```{r, results = 'asis', message = FALSE}
pressure %>% qvis(~temperature, ~pressure) %>% layer_lines()
pressure %>% ggvis(~temperature, ~pressure) %>%
  layer_points() %>% 
  layer_lines()
```

## Histograms

We'll use the built-in `faithful` data set for these examples:

```{r}
# The first few rows
head(faithful)
```


Basic histogram:

```{r, results = 'asis', message = FALSE}
faithful %>% qvis(~eruptions)
faithful %>% ggvis(~eruptions) %>% layer_histograms()
```

Modify the fill color and binwidth, and add titles for the axes, since the automatic titles aren't very informative:

```{r, results = 'asis', message = FALSE}
faithful %>% ggvis(~eruptions, fill := "#fff8dc") %>%
  layer_histograms(binwidth = 0.25) %>%
  set_guide_axis("x", title = "eruptions") %>%
  set_guide_axis("y", title = "count")
```

## Regression lines

Adding a smooth line

```{r, results = 'asis', message = FALSE}
mtcars %>% qvis(~wt, ~mpg) %>% layer_smooths()

mtcars %>% 
  ggvis(~wt, ~mpg) %>%
  layer_points() %>%
  layer_smooths()
```

With a linear model, and 95% confidence interval for the model:

```{r, results = 'asis', message = FALSE}
mtcars %>% qvis(~wt, ~mpg) %>% layer_model_predictions(model = "lm", se = TRUE)
mtcars %>% 
  ggvis(~wt, ~mpg) %>%
  layer_points() %>%
  layer_model_predictions(model = "lm", se = TRUE)
```

## Scatter plots with grouping

Coloring points by a variable:

```{r, results = 'asis', message = FALSE}
mtcars %>% ggvis(~wt, ~mpg) %>% layer_points(fill = ~factor(cyl))
```

Coloring points, and adding a smoother for each group. The grouping variable (which is applied before the `transform_smooth` is calculated) must be specified with `group_by()`:

```{r, results = 'asis', message = FALSE}
mtcars %>% ggvis(~wt, ~mpg) %>% 
  layer_points(fill = ~factor(cyl)) %>% 
  group_by(cyl) %>% 
  layer_model_predictions(model = "lm")
```

# Interactive plots

The following plots won't display correctly in this online document, but they will display correctly when run from R.
